Comparative vaccine effectiveness of ChAdOx1 versus BNT162b2 in Health and Social Care workers in England

### 2021-09-01 analysis/report/draft-manuscript.Rmd
temporarily remove citations

---
title: "Comparative vaccine effectiveness (ChAdOx1 versus BNT162b2) in Health and Social Care workers"
output:
  html_document: 
    keep_md: yes
  word_document: default
  bookdown::html_document2:
    number_sections: false
    toc: false
bibliography: references.bib
csl: nature.csl # from https://raw.githubusercontent.com/citation-style-language/styles/master/nature.csl
link-citations: yes
zotero: true
---

@@

The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease. Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine @polack2020 (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine @voysey2021 (*ChAdOx1*). Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials. To date, there have been no RCTs that have directly compared the BNT162b2 and ChAdOx1 vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same population. The concurrent roll-out of these vaccines across the UK, combined with the country's well-developed electronic health record infrastructure, provides a rare opportunity to emulate such a trial using observational data.

COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease . Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used @collaborative2021. This study assesses the comparative effectiveness of one dose of ChAdOx1 with one dose of BNT162b2 in HCWs, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.

@@

Three outcomes were defined. Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included. We did not differentiate between symptomatic and asymptomatic infection. A&E attendances were identified using HES emergency care records. COVID-19 hospital admissions were identified using HES in-patient hospital records. Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 ("COVID-19, virus identified") and U07.2 ( "COVID-19, virus not identified") were included @emergenc.

@@

We estimated the relative effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR) models [@cupples1988a; @ngwa2016] with the outcome risk estimated at daily intervals. The effect is permitted to vary by time since vaccination to account for the potential differences in vaccine protection over time between the two brands. The PLR models approximate Cox models with time-varying effects, and enables the estimation of risk-adjusted survival rates for each vaccine type using the parametric g-formula. This is one minus the average of the cumulative incidence for each day of follow-up predicted by the PLR model, under the (counterfactual) assumption that every participant received the BNT162b2 vaccine or that every participant received the ChAdOx1 vaccine. Assuming appropriate parameterisation of risk (i.e., adequate confounder adjustment), this replicates the cumulative incidence that would have been observed in an RCT comparing the two vaccines in the population under consideration. Standard errors for the PLR model were obtained using the HC0 clustered sandwich estimator to account for within-participant clustering. Confidence intervals for the risk-adjusted cumulative incidence were obtained using the delta method (i.e, a first-order Taylor series approximation of the variance).

@@

  riskdifference <- 
    cmlinc  %>%
    group_by(model, model_name, outcome, outcome_descr, tstop) %>%
    arrange(tstop, model, outcome, vax1_az) %>%
    mutate(
      diff = (1-survival) - first((1-survival)),
      diff_format = label_number(0.01, scale=1000)(diff)
    ) %>%
    filter(vax1_az==1)

  plot_riskdiff <-
    ggplot(riskdifference)+
    geom_hline(aes(yintercept=0), colour='black')+
    geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
    # geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_grid(cols=vars(outcome_descr))+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=14),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      sec.axis = dup_axis(name="Lower risk after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
    )+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="Risk difference\nper 1000",
      colour=NULL
    )+
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y.left = element_line(colour = "black"),

@@

A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls [@cabezas2021; @vasileiou2021]. [describe pfizer versus AZ findings]

>>>>

---
title: "Comparative vaccine effectiveness of ChAdOx1 versus BNT162b2 in Health and Social Care workers in England"
output:
  html_document: 
    keep_md: yes
    self_contained: TRUE
  word_document: default
  bookdown::html_document2:
    number_sections: false
    toc: false
#bibliography: references.bib
#csl: nature.csl # from https://raw.githubusercontent.com/citation-style-language/styles/master/nature.csl
#link-citations: yes
#zotero: true
---

@@

The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease. Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine <!--# @polack2020 --> (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine <!--# @voysey2021 --> (*ChAdOx1*). Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials. To date, there have been no RCTs that have directly compared the BNT162b2 and ChAdOx1 vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same population. The concurrent roll-out of these vaccines across the UK, combined with the country's well-developed electronic health record infrastructure, provides a rare opportunity to emulate such a trial using observational data.

COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease . Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used <!--#  @collaborative2021 -->. This study assesses the comparative effectiveness of one dose of ChAdOx1 with one dose of BNT162b2 in HCWs, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.

@@

Three outcomes were defined. Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included. We did not differentiate between symptomatic and asymptomatic infection. A&E attendances were identified using HES emergency care records. COVID-19 hospital admissions were identified using HES in-patient hospital records. Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 ("COVID-19, virus identified") and U07.2 ( "COVID-19, virus not identified") were included <!--# @emergenc --> .

@@

We estimated the relative effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR) models <!--# [@cupples1988a; @ngwa2016] --> with the outcome risk estimated at daily intervals. The effect is permitted to vary by time since vaccination to account for the potential differences in vaccine protection over time between the two brands. The PLR models approximate Cox models with time-varying effects, and enables the estimation of risk-adjusted survival rates for each vaccine type using the parametric g-formula. This is one minus the average of the cumulative incidence for each day of follow-up predicted by the PLR model, under the (counterfactual) assumption that every participant received the BNT162b2 vaccine or that every participant received the ChAdOx1 vaccine. Assuming appropriate parameterisation of risk (i.e., adequate confounder adjustment), this replicates the cumulative incidence that would have been observed in an RCT comparing the two vaccines in the population under consideration. Standard errors for the PLR model were obtained using the HC0 clustered sandwich estimator to account for within-participant clustering. Confidence intervals for the risk-adjusted cumulative incidence were obtained using the delta method (i.e, a first-order Taylor series approximation of the variance).

@@

  riskdifference <- 
    cmlinc  %>%
    group_by(model, model_name, outcome, outcome_descr, tstop) %>%
    arrange(tstop, model, outcome, vax1_az) %>%
    mutate(
      diff = (1-survival) - first((1-survival)),
      diff_format = label_number(0.01, scale=1000)(diff)
    ) %>%
    filter(vax1_az==1) %>%
    left_join(
      hazard %>% select(outcome, model, tstop, hr, hr.ll, hr.ul), 
      by=c("outcome", "model", "tstop")
    ) %>% 
    mutate(
      diff.ll = (1-survival) - (1-survival)*hr.ll,
      diff.ul = (1-survival) + (1-survival)*hr.ul
    ) #-(surv-lag(surv))/lag(surv)

  plot_riskdiff <-
    ggplot(riskdifference)+
    geom_hline(aes(yintercept=0), colour='black')+
    geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_grid(cols=vars(outcome_descr))+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=14),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      sec.axis = dup_axis(name="Lower risk after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
    )+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="Risk difference\nper 1000",
      colour=NULL
    )+
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y.left = element_line(colour = "black"),

@@

A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls <!--# [@cabezas2021; @vasileiou2021]  -->. [describe pfizer versus AZ findings]
